Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-12-04 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Fixed syntax errors

This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges. There are four fundamental steps in the analysis of whole-genome bisulfite-sequencing data:

@@

Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq. There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis.

This chapter focuses on steps 1-3 and concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data. The following chapter addresses step 4.

In this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol. Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.

@@

#files <- paste0("../M-bias/", list.files("../M-bias/"))
# ADS
files <- "../M-bias/ADS.M-bias.txt"
sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "ADS") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: ADS") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

```{r M-bias_spike, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:E18VA_M-bias", fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8, dev = "pdf"}
#files <- paste0("../M-bias/", list.files("../M-bias/"))
# E18VA
files <- "../M-bias/E18VA.M-bias.txt"
sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "E18VA") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt\{methclone\} & \cite\{Li:2014ei\}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf\{TODO: Try installing. Suspect you can't get actual methylation calls.\} \\
\texttt\{methpat\}   & \url\{https://github.com/bjpop/methpat\} & Output of \texttt\{bismark\_methylation\_extractor\} & Yes (amplicons)                  &                                                                                  \\
\texttt\{DMEAS\}     & \cite\{He:2013cj\}                       & Output of \texttt\{bismark\_methylation\_extractor\} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabular}
\caption{My caption}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

>>>>

This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges. I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol. Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.

There are four fundamental steps in the analysis of whole-genome bisulfite-sequencing data:

@@

Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq. There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis. This chapter focuses on steps 1-3, with step 4 addressed in Chapter \ref{chap:downstream_analyses}.

The chapter concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data.

@@

#files <- paste0("../M-bias/", list.files("../M-bias/"))
# ADS
files <- "../data/M-bias/ADS.M-bias.txt"
sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "ADS") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: ADS") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

```{r M-bias_spike, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:E18VA_M-bias", fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8, dev = "pdf"}
#files <- paste0("../M-bias/", list.files("../M-bias/"))
# E18VA
files <- "../data/M-bias/E18VA.M-bias.txt"
sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "E18VA") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt{methclone} & \cite{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing. Suspect you can't get actual methylation calls.} \\
\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
\texttt{DMEAS}     & \cite{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabular}
\caption{My caption}
\label{tab:other_m-tuple_methylation_callers}
\end{table}
